The evolution of gene expression in the term placenta of viviparous mammals

### 2016-03-25 manuscript/manuscript.Rnw
add FDR cutoff and ordering by FC for genes

placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5) {
    temp.table <- 
        lineage.table[analysis.name==analysis &
                      p.value <= 0.05][order(p.value),
                                       list(human_name,p.value,fdr,fc),]
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    setkey(temp.table,"Human Symbol")
    if (output) {
        display.table <- temp.table[order(`p (t-test)`),][1:members,]

@@

\begin{table}
  \footnotesize
<<human.lineage.specific,echo=FALSE,results="asis">>=
hum.all.lin.spec <-
    placenta.lineage.specific.table("hum.all",
                                    lineage.table=human.mouse.lineage.specific.analysis,
                                    members=40)
setkey(hum.all.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanationtypeiii{\human}{all other analyzed species}}
\label{tab:human_lineage_specific}
\end{table}

@@

\begin{table}
  \footnotesize
<<mouse.lineage.specific,echo=FALSE,results="asis">>=
mou.all.lin.spec <-
    placenta.lineage.specific.table("mou.all",
                                    lineage.table=human.mouse.lineage.specific.analysis,
                                    members=40
                                    )
setkey(mou.all.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanationtypeiii{\mouse}{all other analyzed species}}
\label{tab:mouse_lineage_specific}
\end{table}

Using a mixed-effects linear model to account for repeated testing of
\human and \mouse, we identified multiple genes which are
differentially expressed in the \mouse lineage in comparison to all
other lineages (\cref{tab:mouse_lineage_specific}). Notable
mouse-specific increases in expression include include genes like
\textit{Slc38a4} which are involved in nutrient transport and are
implicated in fetal growth restriction
\cite{Sharma.ea2016:Tpbpa-Cre-mediateddeletionofTFAP2C}.

>>>>

placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5,orderbyfc=FALSE,fdr.cutoff=1) {
    temp.table <- 
        lineage.table[analysis.name==analysis &
                      p.value <= 0.05][order(p.value),
                                       list(human_name,p.value,fdr,fc),]
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    setkey(temp.table,"Human Symbol")
    if (output) {
        display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
        if (orderbyfc) {
            display.table <- temp.table[order(-abs(`Fold Change`)),]
        }
        display.table <- display.table[1:members,]

@@

\begin{table}
  \footnotesize
<<human.lineage.specific,echo=FALSE,results="asis">>=
hum.all.lin.spec <-
    placenta.lineage.specific.table("hum.all",
                                    lineage.table=human.mouse.lineage.specific.analysis,
                                    members=40,
                                    orderbyfc=TRUE,
                                    fdr.cutoff=0.001
                                    )
setkey(hum.all.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanationtypeiii{\human}{all other analyzed species}}
\label{tab:human_lineage_specific}
\end{table}

@@

\begin{table}
  \footnotesize
<<mouse.lineage.specific,echo=FALSE,results="asis">>=
mou.all.lin.spec <-
    placenta.lineage.specific.table("mou.all",
                                    lineage.table=human.mouse.lineage.specific.analysis,
                                    members=40,
                                    orderbyfc=TRUE,
                                    fdr.cutoff=0.001
                                    )
setkey(mou.all.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanationtypeiii{\mouse}{all other analyzed species}}
\label{tab:mouse_lineage_specific}
\end{table}

Using a mixed-effects linear model to account for repeated testing of
\human and \mouse, we identified multiple genes which are
differentially expressed in the \mouse lineage in comparison to all
other lineages (\cref{tab:mouse_lineage_specific}). \mouse
\textit{Gkn2} (FDR=\Sexpr{mou.all.lin.spec["GKN2",`FDR`]},
FC=\Sexpr{mou.all.lin.spec["GKN2",`Fold Change`]}) has been shown to
have a pattern of expression distinct from the \human pattern, and has
been suggested to be important for placental development
\cite{Fahlbusch.ea2015:Trophoblastexpressiondynamicsof}. Notable
mouse-specific increases in expression include include genes like
\textit{Slc38a4} (FDR=\Sexpr{mou.all.lin.spec["SLC38A4",`FDR`]},
FC=\Sexpr{mou.all.lin.spec["SLC38A4",`Fold Change`]}) and \textit{Afp}
(FDR=\Sexpr{mou.all.lin.spec["AFP",`FDR`]},
FC=\Sexpr{mou.all.lin.spec["AFP",`Fold Change`]}) which are implicated
in fetal growth restriction
\cite{Sharma.ea2016:Tpbpa-Cre-mediateddeletionofTFAP2C} and placenta
abnormalities in humans
\cite{Waller.ea1996:associationbetweenmaternalserum}, and
\textit{Cnx2} (FDR=\Sexpr{mou.all.lin.spec["CNX2",`FDR`]},
FC=\Sexpr{mou.all.lin.spec["CNX2",`Fold Change`]}) which is critical
for the exchange of metabolites in the placenta labyrinth
\cite{Gabriel.ea1998:Transplacentaluptakeofglucose}.
